Bayesian exponential random graph modeling of KWL Burial network

### 2020-10-14 analysis/paper/paper.Rmd
edit Intro

<!--burial study- network analysis-->
Burials provide valuable information about past human behavior that can reflect not only the personal life and status of the deceased, but also the expressions of their descendants through grave goods and arrangements [@Pearson1999]. Mortuary practices such as burial forms, grave goods, and related ritual practices could represent the social relations between individuals and allow archaeologists to infer social structures. This enables the study of social differences based on burials to answer questions related to social inequality and complexity [@Ehrhardt2009; @Gamble2002; @Janes2013; @Johansen2015]. For examples, the use of burial goods can reflect social practices in a broader of cultural contexts to present personal wealth, social roles, or a traditional custom. Burials with grave goods in common often indicates underlying social relation connected through trade, exchange, and gifting [@Coward2013, pp. 252]. Identifying the structure of relationships among burials allows a further inferences on a structure of a social organization. A common way to study connections between individuals is network analysis, which has been applied increasingly to archaeological studies recently [@Mills2017]. Networks are represented by graphs that consists of a set of actors, such as people, groups, or organizations, and a set of links connecting them. Technically speaking, actors are "nodes" connected by multiple "edges", representing relationships, between nodes. The basic unit of relationship is called dyad, defined as a connection between two distinct actors [@Lee2016]. In archaeology, relationships can be viewed as flows of resources that reflect relations of control, dependence, and cooperation, reflecting social relations in societies [@Marin2011]. With network analysis, we can quantify, visualize, and statistically test hypotheses of the relationships between actors to examine the structure of network and major factors that affects the structure [@Freeman2004; @Salvini2010]. The application in archaeology is widely used for [@Coward2013; @Mizoguchi2013; @Potgieter2008]

<!--archaeology case-->
The archaeological sites in northeastern Taiwan show evidence of use of imported prestige goods in burials and accumulation of them in residential area around the same time as European contact in the 17th century [@Chen2007], which might indicate the pursuit of prestige or wealth in the society. Those prestige goods include imported ornaments and porcelains that entered northeastern Taiwan through regional exchange networks involved the Europeans, Han Chinese traders, and Indigenous groups. In addition to prestige, European colonial powers could have provided military support to indigenous allies. For example, @Voss2005's study of a colonial outpost in California shows that colonial residents have different identities and higher social status than other local indigenous people. A general model to summarize these case studies might be that the support of a colonial power combined with high local values for imported goods, might lead to competition among individuals. In this study, we will use this model of competition to explain the changes in social inequality at Kiwulan, an Iron Age site in northesatern Taiwan, which covers the presence of the Europeans, including the Spanish and the Dutch, in the 17th century and the Chinese in the 19th century.

@@

Using the R software, we apply Bayesian SNA to burial data from Kiwulan, an Iron Age site in northeast Taiwan. We view individual burials as actors or nodes in the social network with ties drawn between burials when the presence of the same variables that can represent the same "community" or social group. We test hypotheses about the distribution of influence in the network to identify network or corporate mode conditions. We ask the questions: 1. Do the observed burial data indicate a more clustered network than a distribution of random networks with similar qualities? 2. Whether European colonial activities in 17 century Taiwan resulted in the emergence of social inequality in an indigenous society can be detected by burials networks. 3. How the relations among burials may affect the quantity of burial goods? We expect the network to present a higher degree of clustering (e.g. the number of completed triangles in the network) in the European contact period that might hint at social heterogeneity during this period. This study helps to expand the use of burials in understanding the indirect effects of a colonial presence on indigenous groups. A Bayesian approach enables the efficient quantification for uncertainty, parametrization, and model evaluation of social network metrics.

@@

<!--archaeological setting-->
We analyzed burial data collected from the upper component of Kiwulan site, an Iron Age settlement site (1350-1850 AD) located in northeastern Taiwan, which experienced the European colonial impacts in the early 17th century and the large wave of Chinese immigrants in the 19th century [@Chen2007]. The excavation revealed abundant pottery sherds, imported ceramics and stonewares, wooden artifacts, stone tools, metal artifacts, imported glass beads and agates beads, and pipes [@Chen2007]. In addition to these artifacts, 90 burials, hundreds of middens, storage pits, and postholes with in-situ posts were also excavated. The burials are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison. Those burials are oriented in an east-west direction on the north side of the house structures indicated by postholes and `in-situ` wooden posts, indicating a well organized space arrangement.

Previous studies on burials suggest a presence of uneven distribution of prestige goods across burials. However, there is a debate over whether the differentiation between burials hints an economic inequality supporting an egalitarian society [@Hsieh2012], or political inequality indicating a stratified society [@Cheng2008]. @Cheng2008 interprets the unequal distribution of prestige materials, especially the gold-foil beads between burials, as evidence for hierarchy. However, @Hsieh2012 suggests a relatively egalitarian society based on the comparative analysis of frequency and proportion of all burial goods. She found that the burials with rich goods were usually associated with elders, which might indicate achieved, rather than inherited, status. The earliest record of the use of prestige goods in burials can be traced back around the European arrival according to historical records from the Spanish and the Chinese [@Borao2009; @LiandWu2006]. For example, the Spanish document by a Spanish Priest described that the Spanish soldier used agate beads to exchange materials from local indigenous people since they were valuable in their culture [@LiandWu2006]. In addition, there was a custom that people would bury imported materials with deceased to display their status and influence [@LiandWu2006]. Despite we observed the economic differentiation in burial goods and the high value of imported goods according to historical documents, the degree of inequality over time, and which variable decides the changes remain unclear. Analyzing and comparing burial data from different phases using network analysis approaches will give an insight into the degree and changes in social inequality. This is important to understand the social reactions of Kiwulan residents in the context of the foreign contact.

>>>>

<!--burial study- network analysis-->
Burials contains valuable information about individual mortuary practices and inherited structure of a society as a whole. With the development of network science and graph theory, it is promising to study relation among burials and explore its social implication [@Sosna2013]. Mortuary practices such as burial forms, grave goods, and related ritual practices could represent the social relations between individuals and allow archaeologists to infer social structures. This enables the study of social differences based on burials to answer questions related to social inequality and complexity [@Ehrhardt2009; @Gamble2002; @Janes2013; @Johansen2015]. For examples, the use of burial goods can reflect social practices in a broader of cultural contexts to present personal wealth, social roles, or a traditional custom. Burials with grave goods in common often indicates underlying social relation connected through trade, exchange, and gifting [@Coward2013, pp. 252]. Identifying the structure of relationships among burials allows a further inferences on a structure of a social organization. A common way to study connections between individuals is network analysis, which has been applied increasingly to archaeological studies recently [@Mills2017]. Networks are represented by graphs that consists of a set of actors, such as people, groups, or organizations, and a set of links connecting them. Technically speaking, actors are "nodes" connected by multiple "edges", representing relationships, between nodes. The basic unit of relationship is called dyad, defined as a connection between two distinct actors [@Lee2016]. In archaeology, relationships can be viewed as flows of resources that reflect relations of control, dependence, and cooperation, reflecting social relations in societies [@Marin2011]. With network analysis, we can quantify, visualize, and statistically test hypotheses of the relationships between actors to examine the structure of network and major factors that affects the structure [@Freeman2004; @Salvini2010]. The application in archaeology is widely used for [@Coward2013; @Mizoguchi2013; @Potgieter2008]

<!--archaeology case-->
The archaeological sites in northeastern Taiwan show evidence of use of imported prestige goods in burials and accumulation of them in residential area around the same time as European contact in the 17th century [@Chen2007], which might indicate the pursuit of prestige or wealth in the society. Those prestige goods include imported ornaments and porcelains that entered northeastern Taiwan through regional exchange networks involved the Europeans, Han Chinese traders, and Indigenous groups. In addition to prestige, European colonial powers could have provided military support to Indigenous allies. For example, @Voss2005's study of a colonial outpost in California shows that colonial residents have different identities and higher social status than other local Indigenous people. A general model to summarize these case studies might be that the support of a colonial power combined with high local values for imported goods, might lead to competition among individuals. In this study, we will use this model of competition to explain the changes in social inequality at Kiwulan, an Iron Age site in northesatern Taiwan, which covers the presence of the Europeans, including the Spanish and the Dutch, in the 17th century and the Chinese in the 19th century.

@@

Using the R software, we apply Bayesian SNA to burial data from Kiwulan, an Iron Age site in northeast Taiwan. We view individual burials as actors or nodes in the social network with ties drawn between burials when the presence of the same variables that can represent the same "community" or social group. We test hypotheses about the distribution of influence in the network to identify network or corporate mode conditions. We ask the questions: 1. Do the observed burial data indicate a more clustered network than a distribution of random networks with similar qualities? 2. Whether European colonial activities in 17 century Taiwan resulted in the emergence of social inequality in an Indigenous society can be detected by burials networks. 3. How the relations among burials may affect the quantity of burial goods? We expect the network to present a higher degree of clustering (e.g. the number of completed triangles in the network) in the European contact period that might hint at social heterogeneity during this period. This study helps to expand the use of burials in understanding the indirect effects of a colonial presence on Indigenous groups. A Bayesian approach enables the efficient quantification for uncertainty, parametrization, and model evaluation of social network metrics.

@@

<!--archaeological setting-->
We analyzed burial data collected from the upper component of Kiwulan site, an Iron Age settlement site (1350-1850 AD) located in northeastern Taiwan, which experienced the European colonial impacts in the early 17th century and the large wave of Chinese immigrants in the 19th century [@Chen2007]. The excavation revealed abundant pottery sherds, imported ceramics and stonewares, wooden artifacts, stone tools, metal artifacts, imported glass beads and agates beads, and pipes [@Chen2007]. In addition to these artifacts, 90 burials, hundreds of middens, storage pits, and postholes with in-situ posts were also excavated. The burials are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison. Those burials are oriented in an east-west direction on the north side of the house structures indicated by post-holes and `in-situ` wooden posts, indicating a well organized space arrangement.

Previous studies on burials suggest a presence of uneven distribution of prestige goods across burials. However, there is a debate over whether the differentiation between burials hints an economic inequality supporting an egalitarian society [@Hsieh2012], or political inequality indicating a stratified society [@Cheng2008]. @Cheng2008 interprets the unequal distribution of prestige materials, especially the gold-foil beads between burials, as evidence for hierarchy. However, @Hsieh2012 suggests a relatively egalitarian society based on the comparative analysis of frequency and proportion of all burial goods. She found that the burials with rich goods were usually associated with elders, which might indicate achieved, rather than inherited, status. The earliest record of the use of prestige goods in burials can be traced back around the European arrival according to historical records from the Spanish and the Chinese [@Borao2009; @LiandWu2006]. For example, the Spanish document by a Spanish Priest described that the Spanish soldier used agate beads to exchange materials from local Indigenous people since they were valuable in their culture [@LiandWu2006]. In addition, there was a custom that people would bury imported materials with deceased to display their status and influence [@LiandWu2006]. Despite we observed the economic differentiation in burial goods and the high value of imported goods according to historical documents, the degree of inequality over time, and which variable decides the changes remain unclear. Analyzing and comparing burial data from different phases using network analysis approaches will give an insight into the degree and changes in social inequality. This is important to understand the social reactions of Kiwulan residents in the context of the foreign contact.
